---
title: "Week 1 Case: TSLA and APTV"
author: "Your Names"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
library(rethinking)
library(tidyverse)
library(kable)
library(kableExtra)
```

## The assignment

Many know of Elon Musk's Tesla (TLSA), few might know much about a major supplier called Aptiv plc (APTV) with slogan:

> Making Mobility Greener

We want to know if APTV's stock price movements are influenced by TSLA. The analysis will itself move in three steps:

1. What is the probability of an up movement in APTV, and how good a model is that?

2. What is the probability of an up movement in TSLA, and again, how good a model is that?

3. If TSLA's stock drops, what are the odds that APTV's stock will also drop?

We collect three months of daily stock price adjusted closing price data from finance-yahoo, process the data into series of up and down movements, here is a wrangle we can use.

```{r}
options(digits = 2, scipen = 99999 )
price <- read_csv("tsla-aptv.csv")
r_all <- tibble(r_tsla = diff(log(price$tsla)), r_aptv = diff(log(price$aptv)) )
r_all <- r_all %>% mutate( move_tsla = ifelse( r_tsla > 0, "UP", "DOWN"), move_aptv = ifelse( r_aptv <= 0, "UP", "DOWN"))
summary( price )
summary( r_all )

```

## APTV movements


## TLSA movements


## Do TSLA stock prices influence APTV?

We have a binary world here: stockprices go up and they go down. 

> 3. If TSLA's stock drops, what are the odds that APTV's stock will also drop?

We will need to construct the Odds Ratio that, given TSLA stock prices go down, what is probability that APTV prices will also go down relative to the alternative when TSLA prices fall. The alternative is that APTV prices rise.


### Two ways tables

```{r}
r_all %>% 
  with( addmargins( table( move_aptv, move_tsla)) )

r_all %>% 
  with(addmargins( proportions(table( move_aptv, move_tsla))))

```

Just like the land-water problem on Earth-Mars, we can use these frequencies and relative frequencies to answer our questions about the joint effect of TSLA stock price movements on APTV.

```{r}
pr_APTV-down_TSLA-down <- 42
pr_APTV-up_TSLA-down <- 42
odds_APTV-down_TSLA-down <- 42
```

### Comments on results

- ONE

- TWO



## Chapter 2 -- Small worlds and large worlds

### 2M3: My Favorite Martian

Here's the data from the problem:

$$
\begin{align}
Pr(land \mid Earth) &= 1 − 0.7 \\
Pr(land \mid Mars) &= 1
\end{align}
$$

Are there not equal prior expectations of each globe?

$$
\begin{align}
Pr(Earth) &= 0.5 \\
Pr(Mars) &= 0.5
\end{align}
$$

Now the piece de resistance, Bayes,

$$
Pr(Earth \mid land) = \frac{Pr(land \mid Earth)Pr(Earth)}{Pr(land)}
$$
We need to calculate the $Pr(land)$. If we set this up as a contingency table we might be able to see that this probability is the expected value of seeing land whether on Mars or on Earth. Those *ors* mean we need to add the probabilities of Earth and land along with the probability of Mars and land by the probabilities respectively of seeing Earth and Mars. Thus we have

$$
Pr(land) = Pr(Earth)Pr(Earth\,\,and\,\,land) + Pr(Mars)Pr(Mars\,\,and\,\,land )
$$
This thinking is the core of all of the rest of the course!

We code this in R as

```{r martians}
pr_earth_land <- 0.42 # It is not 42! Let R do the arithmetic
```

The answer is `r pr_earth_land`.

### 2M2: We know a little something

We start with this model and the improper prior (does not add to one like a good probability distribution should).

```{r week-start}
p_grid <- seq( from=0 , to=1 , length.out=100 )
likelihood <- dbinom( 3 , size=3 , prob=p_grid )
prior <- 1
posterior <- likelihood * prior
posterior <- posterior / sum(posterior) # standardize
plot( posterior ~ p_grid , type="l" )
```

In this code chunk we insert a new prior `prior <- ifelse( p_grid < 0.5 , 0 , 1 ) # new prior` in place of the improper `1` and rerun.

```{r redo}

```

Next we try adding more land than water in this chunk with this data:

$$
L, W, W, L, W, W, W
$$

```{r more-land}
# moreland
```

The upshot?

- Point one

- Point two

- Kai toi loipa, usw, etc.

## Chapter 3 -- Sampling the imaginary

### Start up with 3M1-2

Here we input the new number of tosses and results. We just change the data in this existing model.

```{r 3M1}
p_grid <- seq( from=0 , to=1 , length.out=1000 ) # hypptheses
prior <- rep( 1 , 1000 )
tosses <- 9 # data
water <- 6 # data
likelihood <- dbinom( water , size=tosses , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
plot( posterior ~ p_grid , type="l" )
```

### Next 3M2-3 and a predictive check

Draw samples first, then check the results with a density plot and a 90\% HPDI (Highest Posterior Density Interval) from the `rethinking` package.

```{r sampling-checking}
library(rethinking)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
dens( samples )
HPDI( samples , prob=0.9 ) #90% HPDI
```

Interpretation:

- The HPDI indicates...

- The density function shows that...

We now run a predictive check on the posterior distribution. This routine ...

1. Sample ...

2. Simulate ...

3. Compare simulation with the data that `w==8`. Calculate the proportions.

```{r pp-check}
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
w <- rbinom( 1e4 , size=tosses , prob=samples )
# Look this sum() / 1e4 ratio in the chapter and calculate here. Use the simplehist() of w to plot the results

```

What does this mean?

- A good model?

- Compatibility?

### Now to change up the priors 3M5

Start with the alternate priors from 2M2.

```{r 2M2-priors}
# alternate priors
```

Run the `HPDI()` and simulate `w <- rbinom()` and show `simplehist()`. 

```{r redo-new-prior}
# check again
```

Some questions to ponder.

1. We don't trust the data? Is that what the new prior is saying? 

2. What about the observed value off 8. What happened to the predictive distribution?

3. What does this mean for building models? Communicating with those who would consume your analysis?

